Effective Inclusion of Electronic Polarization to a Negatively Charged Membrane

### 2019-03-07 Manuscript/manuscript.tex
Added some missing literature and included exceprts from recent literature to be discussed in the paper in the section with Ca2+ which is still to be heavily edited.

Further details about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS
were obtained by counting contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18}. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
approximately 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
This corroborates the intrinsically higher affinity of PS lipids to calcium cations compared to neutral PC lipids. 
POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral and negatively charged membranes. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present, 
however, they are always accompanied by interactions with phosphate groups. 
%but a very small negligable amount much below 1\%. 
\todo{Also COO- vs. phosphate binding should be discussed and the results should be
  discussed with respect to the literature: https://doi.org/10.1016/j.bpj.2018.08.044 and
  Melcrova et al. Scientific Reports | 6:38035 | DOI: 10.1038/srep38035 }
%
\todo{EDIT: Taken from NMRlipids4 as an interesting information, to be edited.}
The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%
CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
the same force field without the NBfix parameters together with the NMR chemical shifts and
REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

@@

Provide density plots and show, where and how the cations bind to the membrane.

>>>>

% Description of no. conctacts counting
Further details about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS
were obtained by counting contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18}. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

% Populations from contacts
The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
approximately 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
This corroborates the intrinsically higher affinity of PS lipids to calcium cations compared to neutral PC lipids. 
% Where Calcium interacts
POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral \cite{melcr18} and negatively charged membranes. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present, 
however, they are always accompanied by interactions with phosphate groups. 
%but a very small negligable amount much below 1\%.

@@

FOLLOWING: lit. search on recent literature, which shall be discussed in this work.

\todo{EDIT: Taken from NMRlipids4 as an interesting information, to be edited.}
The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%
CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
the same force field without the NBfix parameters together with the NMR chemical shifts and
REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

% Details of interactions between calcium and phospholipids (PC and PS) .
% Major conclusions correct, only the model (Berger) over-exaggerates certain features. 
From abstract of \cite{melcrova16}://
Namely, time-resolved fluorescent spectroscopy of lipid vesicles 
and vibrational sum frequency spectroscopy of lipid
monolayers are used to characterize local binding sites of calcium in zwitterionic and anionic model
lipid assemblies, while dynamic light scattering and zeta potential measurements are employed for
macroscopic characterization of lipid vesicles in calcium-containing environments. To gain additional
atomic-level information, the experiments are complemented by molecular simulations that utilize an
accurate force field for calcium ions with scaled charges effectively accounting for electronic polarization
effects. We demonstrate that lipid membranes have substantial calcium-binding capacity, with several
types of binding sites present. Significantly, the binding mode depends on calcium concentration with
important implications for calcium buffering, synaptic plasticity, and protein-membrane association.

And From conclustions of \cite{melcrova16}://
Calcium ion binding sites are heterogeneous both from the point of view of binding affinity and their positioning
in the membrane. The character of the calcium binding varies with calcium concentration; this issue is of par-
ticular importance as significant concentration spikes of calcium ions occur along calcium signaling pathways.
The present results support the conjecture that lipid membranes, in particular the negatively charged inner leaflet
of the plasma membrane, can act as effective calcium buffers upon calcium ions entering the cytosol. Of equal
importance is the fact that the strong binding of this ion significantly alters the membranes by means of reduction
of their hydration, lipid mobility, and lateral inter-lipid distance. Such local conformational membrane remod-
eling may play a significant role in modulation of lipid-protein interactions as well as membrane-membrane
interactions. Overall, the phenomena related to calcium ions-membrane interactions demonstrated here point to
their diverse roles in biological systems.

% Ca2+ induces much slower dynamics of lipids when PS is present,
% this does not happen with PC
From Ref.~\cite{valentine18}://
To summarize, calcium binding to PS lipids slows the dynamics at the ester position and introduces more heterogeneous molecular environments. 
The slowdown cannot be attributed to complete ester dehydration or decreased hydrogen bonding to water, so we interpret these results as evidence that the lipid headgroups adopt more rigid but disordered conformations on the molecular level. 
The decrease in dynamics observed experimentally indicates that the network of hydrogen bonds between lipids and water at the polar-nonpolar interface becomes slower, as it was shown by Karathanou and Bondar (34). 
This is bolstered by recent NMR work showing that binding to Ca2+ induces two different rigid conformations in the glycerol backbone of lipid headgroups (18).
We have investigated the molecular origin of the rigidification and decreased lateral diffusion, finding that calcium binding does not dehydrate the ester region, but even with only 10\% PS lipids, the local dynamics slow around the esters. The slower local dynamics provide a molecular explanation for the Ca2+-dependent decreases in fluidity observed on longer length and timescales.
There is considerable room for further improvement in the force fields for ion-membrane interactions. Particularly, small differences in short-range interactions may determine the local geometries (34
). Most of our conclusions are, however, based on long-range effects (i.e., electric field generated by the ion at significant distances from the carbonyl groups), which are less sensitive to the details of the force field.

% 2 major head group configurations of PS in 30-70 PC-PS mix -- paper 1
From Ref.~\cite{boettcher11}://
 In this study, combining magic angle spinning (MAS) solid-state NMR (SSNMR) measurements of isotopically labeled serine headgroups in mixed lipid bilayers with molecular dynamics (MD) simulations of PS lipid bilayers in the presence of different counterions, we provide site-resolved insights into the effects of Ca2+ on the structure and dynamics of lipid bilayers. Ca2+-induced conformational changes of PS in mixed bilayers are observed in both liposomes and Nanodiscs, a nanoscale membrane mimetic of bilayer patches. Site-resolved multidimensional correlation SSNMR spectra of bilayers containing 13C,15N-labeled PS demonstrate that Ca2+ ions promote two major PS headgroup conformations, which are well resolved in two-dimensional 13C-13C, 15N-13C, and 31P-13C spectra. The results of MD simulations performed on PS lipid bilayers in the presence or absence of Ca2+ provide an atomic view of the conformational effects underlying the observed spectra.
but the NMR ...
 Spectra were acquired of 70\% POPS* and 30\% POPC Nanodiscs with 2 mM Ca2+ using an MAS rate of 10 kHz on a 600 MHz (1H frequency) spectrometer
and but the MD sim was done only for pure POPS bilayer and no other counterions but Ca2+ with CaCl2 concentration.

% 2 major head group configurations of PS in 30-70 PC-PS mix -- paper 2
% extended analysis and insight. 
From Ref.~\cite{hallock18}://
(conclusions) We investigated the atomistic details of how membranes containing anionic PS lipids are shaped into their functional state by Ca2+ ions. The lipid phosphate groups and the captured ions are made readily accessible to interact with proteins. Using extensive MD simulations with a lipid-accelerated membrane representation, we demonstrated that PS lipid nanoclusters are formed by two distinct PS headgroup conformations that are induced by Ca2+ ions, which bind to the lipids in a specific manner. Importantly, detailed characterization of MD-derived lipid nanoclusters and geometries of individual PS lipids agree with SSNMR measurements on two different spatial scales. (1) The chemical shifts show the accuracy of capturing the chemical environment around a single lipid in either of the two long-lived conformations, and (2) the Ca2+-coordinated PS lipid tetramer is a basic structural unit that is involved in membrane sculpturing. By studying the effects of ions on membranes containing anionic lipids, we have characterized a baseline behavior and structures of the lipids. These results will be important in interpreting the role of lipids in future protein–membrane studies, and they demonstrate that MD HMMM simulations combined and validated by SSNMR methodologies are effective at capturing high-resolution signature lipid structures that are stabilized by membrane-active agents in complex membrane environments.
but ... again, the experiment with nanodiscs (and also the bilayer simulation with HMMM) 
uses 70:30 PS:PC ratio, which induces PS clustering.
 -> hence the markable difference between their figure 3 and our figure \ref{fig:cacl_complexes}
 - they observe most probable Ca2+ clusters of 2, then 3, 1 and 4 PS together
 - while we see rather only 1PS per ca2+ in Figure \ref{fig:cacl_complexes}
On the other hand, I see correspondence with the observed longer life time ( stronger interaction?)
with \ce{COO-} group in PS in \cite{hallock18} in Fig 4, for both Ca2+ and Na+. 
This is followed by PO4 in both cases. 
In the case of Ca2+, the ratio of the observed life time is also in line with our findings, that 
the interaction with the PO4  and COO- together is of a smaller abundance compared to the interactions of the individual groups themselves, yet comparable to that of PO4. 
Their Figure 5 shows the distributions of O-Ca-Cb-N dihedral angle
revealing separate populations of the PS head groups. 
These two structures are also compared in therm of P-Cg distances, 
which can be measured from NMR REDOR exp. 
It is concluded that since the two distinct populations in P-Cg distance can be represented with the 
dihedral distributions, 
they also likely represent the distinct populations. 
 -> plotting the distribution of O-Ca-Cb-N dihedral angle may be also a good idea for us (SI?)
     -> NMRlipids 4 in SI \cite{nmrlipids_proj4} plots this information: 
           - C36 agress with the distribution in \cite{hallock18}
           - Lipid17 (and Slipids) are almost identical to C36
           - but the structure of the COO- is highly different (dih angle Ca-Cb-Cg-Og)
